A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 3.0)

### 2015-09-24 control.tex
more updates

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskeleton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot (we indicate this with
the subscript $\ox$ in dynamics equations.
We can also separate out the operator dynamics (indicated by
the subscript $\myo$) from the exoskeleton dynamics (indicated by
the subscript $\myx$) if this is more convenient
(we separate them out and ignore them in what immediately follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} + \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
+ \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
\end{equation}
$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
force $\vf_{\ox}$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

@@

A control objective could be to have the exoskeleton to autonomously
apply a desired force.
One could add
\begin{equation}
\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
\end{equation}
to a gravity compensated control or to control including inverse dynamics.

Active force control could be added to improve the performance:
\begin{equation}
\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
\end{equation}

@@

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.
Let's describe a small position offset of the exoskeleton-world
contact points as $\vq_w$. We know that:
\begin{equation}
\dot{\vq}_w = \mJ_w(\vq) \dot{\vq}
\end{equation}
and that
\begin{equation}
\vtau = \mJ_w^{\tr}(\vq) \vf_w 
\end{equation}
This allows us to relate stiffness and damping in the external world coordinates
with the stiffness and damping in exoskeleton joint coordinates.
Here is the relationship for damping:
\begin{equation}
\vtau = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq) \dot{\vq}
\end{equation}
So the damping in joint coordinates is $\vtau = \mK_{\dot{\vq}} \dot{\vq}$ and
therefore
\begin{equation}
\mK_{\dot{\vq}}(\vq) = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq)
\end{equation}
The mapping is the same for stiffnes:
\begin{equation}
\mK_{\vq}(\vq) = \mJ_w^{\tr}(\vq) \mK_{w} \mJ_w(\vq)
\end{equation}
So we should assign the exoskeleton joint position gains to be $\mK_{\vq}(\vq)$ and
the joint velocity gains to be $\mK_{\dot{\vq}}(\vq)$ in equation~\ref{eq:impedance}.

Similarly, we can use exoskeleton-world force control to improve performance,
setting
\begin{equation}
{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
\end{equation}
and add the force error: 
\begin{equation}
\vtau = \vtau_{\xinvdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+ \mJ_w^{\tr}(\vq) \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
\end{equation}

@@

This section describes how to handle underactuation and overactuation using
``floating body dynamics.''
We have been sweeping this
issue under the rug so far. It is now useful to be more explicit.
There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Humanoid robots (and humans) have floating body dynamics.
A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. A selection matrix
$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau = \mS \vu
\end{equation}
The rows in $\mS$ corresponding to the root linear and angular velocities
are all zero.

The second kind of underactuation appears in exoskeletons where we rely on the
operator to actuate a particular joint. For example, many lower body
exoskeletons only actuate the joints in the sagittal plane, leaving lateral
and twist (yaw) joints unactuated or passively actuated.

@@

The underactuation can be handled by reducing the size of the $\vu$ vector
and changing the matrix $\mS$ to reflect the missing actuation.
We can have redundant actuation (more than one actuator across any particular
joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
We can have multijoint actuation, which can be handled in $\vu$ and $\mS$ as well.

The modified dynamics equations are
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:dynamics2}
\end{equation}
Note that $\ddot{\vq}$ and the new actuation vector $\vu$
do not have the same dimensionality. What difference does that make?

@@

Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
no longer works if $\mS$ is not invertable (not square or not full rank).
In the underactuated and overactuated cases $\mS$ is not square. 
In these cases {\bf online optimization}
as described in the next Section is used to ``perform inverse dynamics'' by
finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
and actuator commands $\vu$.

@@

@@

sensitivity to
 modeling error
 sensor error
 unmodeled dynamics

underactuated control problem
 - not fully actuated sagittal only.
 - how necessary is lateral actuation
 - what is the penalty
  weight
  drag of actuator

how handle failure and damage after Oct 1

what about a full model of the system

Oct 5-9

>>>>

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskeleton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot (we indicate this with
the subscript $\ox$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
terms in the dynamics equations).
We can also separate out the operator dynamics (indicated by
the subscript $\myo$ on ICCG terms) from the exoskeleton dynamics (indicated by
the subscript $\myx$ on ICCG terms) if this is more convenient
(we separate them out and ignore operator dynamics in what immediately follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} - \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
- \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
\end{equation}
$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
force $\vf_{\ox}$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

@@

A control objective could be to have the exoskeleton to autonomously
apply a desired force.
One could add
\begin{equation}
\mJ_{\xw}^{\tr}(\vq) {\vf_{\xw}}_d
\end{equation}
to a gravity compensated control or to control including inverse dynamics.

Active force control could be added to improve the performance:
\begin{equation}
\mJ_{\xw}^{\tr}(\vq) \mK_{\vf_{\xw}} (\vf_{\xw} - {\vf_{\xw}}_d)
\end{equation}

@@

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.
Let's describe a small position offset of the exoskeleton-world
contact points as $\vq_w$. We know that:
\begin{equation}
\dot{\vq}_w = \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
and that
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
\end{equation}
This allows us to relate stiffness and damping in the external world coordinates
with the stiffness and damping in exoskeleton joint coordinates.
Here is the relationship for damping:
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
So the damping in joint coordinates is $\vtau_{\myx} = \mK_{\dot{\vq}} \dot{\vq}$ and
therefore
\begin{equation}
\mK_{\dot{\vq}}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq)
\end{equation}
The mapping is the same for stiffnes:
\begin{equation}
\mK_{\vq}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{w} \mJ_{\xw}(\vq)
\end{equation}
So we should assign the exoskeleton joint position gains to be $\mK_{\vq}(\vq)$ and
the joint velocity gains to be $\mK_{\dot{\vq}}(\vq)$ in equation~\ref{eq:impedance}.

Similarly, we can use exoskeleton-world force control to improve performance,
setting
\begin{equation}
{\vf_{\xw}}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
\end{equation}
and add the force error: 
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+ \mJ_{\xw}^{\tr}(\vq) \mK_{\vf_{\xw}} ( \hat{\vf}_{\xw} - {\vf_{\xw}}_d )
\end{equation}

@@

This section describes how to handle underactuation and overactuation using
``floating body dynamics.''
We have been sweeping this
issue under the rug so far. It is now useful to be more explicit.
There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Contacts are made and broken during walking, running, and many other behaviors,
and contacts slip.
Humanoid robots (and humans) have floating body dynamics.
A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. A Jacobian matrix
$\mJ_{\myx}$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau_{\myx} = \mJ_{\myx} \vu_{\myx}
\end{equation}
The rows in $\mS$ corresponding to the root linear and angular velocities
are all zero.
To maintain compatability in our joint representation, we will similarly
map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
\begin{equation}
\vtau_{\myo} = \mJ_{\myo} \vu_{\myo}
\end{equation}

The second kind of underactuation appears in exoskeletons where we rely on the
operator to actuate a particular joint. For example, many lower body
exoskeletons only actuate the joints in the sagittal plane, leaving lateral
and twist (yaw) joints unactuated or passively actuated. 
Many ``exoskeletons'' or orthoses only actuate one joint on each limb (typically
the ankle or knee). In these situations the question arises as to what is
the scope of control: should the device only consider
just the actuated joints,
or should it consider a wider range of operator joints? Many of these devices
do not have enough sensing (such as an IMU) to even estimate the state of other
operator joints. We believe that for high performance the entire state of the
operator must be considered in control, and we describe a way to do it in
Section~\ref{sec:qp}.
However, we are not aware of any system that goes this far in terms of control.

@@

``Missing joint''
underactuation can be handled by reducing the size of the $\vu_{\myx}$ vector
and changing the matrix $\mJ_{\myx}$ to reflect the missing actuation.
We can have redundant actuation (more than one actuator across any particular
joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accomodate extra actuators.
We can have multijoint actuation, which can be handled in $\vu_{\myx}$ 
and $\mJ_{\myx}$ as well.

The modified dynamics equations are
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \mJ_{\myo} \vu_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:odynamics2}
\end{equation}
\begin{equation}
\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} = \mJ_{\myx} \vu_{\myx} + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:xdynamics2}
\end{equation}
\begin{equation}
\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow}
+ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
= \mJ_{\myo} \vu_{\myo} + \mJ_{\myx} \vu_{\myx}
\label{eq:oxdynamics2}
\end{equation}
Note that $\ddot{\vq}$ and the actuation vectors $\vu_{\myo}$ and $\vu_{\myx}$
do not have the same dimensionality. What difference does that make?

@@

Full (operator) body
{\bf online inverse dynamics} as described in Section~\ref{sec:invdyn}
no longer works if $\mJ_{\myx}$ is not invertable (not square or not full rank).
In the underactuated and overactuated cases $\mJ_{\myx}$ is not square. 
For a partial (operator) body
exoskeleton we can rewrite the dynamics in terms of only those
joints the exoskeleton actuates, and invert those dynamics to cancel exoskeleton
effects on those joint. This is what has been done with the BLEEX exoskeleton,
for example.~\cite{}.

However, {\bf online optimization}
as described in the next Section can be used to perform full (operator)
body inverse dynamics by
finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
and actuator commands $\vu_{\myx}$.

@@

Note that the same approach could be used to minimize or control internal forces
such as skeletal forces (forces between bones) at the operator joints to protect
the ankle, hip, and knee joints, for example.
MORE GOES HERE.

@@

sensitivity to
 modeling error
 sensor error
 unmodeled dynamics
* future (after Oct 1) simulation studies

underactuated control problem
 - how necessary is lateral actuation
   * we don't know
 - * what is the penalty
  weight
  drag of actuator
* future (after Oct 1) simulation studies

how handle failure and damage 
* after Oct 1
